Adaptive group-regularized logistic elastic net regression

### 2019-02-05 docs/manuscript_biostatistics.Rnw
adjusted alzheimer data analysis

\documentclass[a4paper,hidelinks]{article}
\usepackage{url,bm,algorithm,algpseudocode,mathtools,bbm,pgfplotstable,bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,rotating,caption,subcaption,natbib,appendix,titlesec,multicol,hyperref,verbatim,bbm,algorithm,algpseudocode,pgfplotstable,threeparttable, booktabs,mathtools,dsfont,parskip}

@@

% settings
\graphicspath{{../figs/}}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Adaptive group-regularized logistic elastic net regression}
\date{}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Carel F.W. Peeters$^{1}$, Aad W. van der Vaart$^{2}$, and \\ Mark A. van de Wiel$^{1,3}$}

@@

@@

	\noindent
	1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health research institute, VU University Medical Center, PO Box 7057, 1007 MB
	Amsterdam, The Netherlands\\
	2. Mathematical Institute, Leiden University, Leiden, The Netherlands \\
	3. MRC Biostatistics Unit, Cambridge, United Kingdom \\

	\begin{abstract}
		{In high-dimensional data settings, additional information on the features is often available. Examples of such external information in omics research are: (a) $p$-values from a previous study, (b) a summary of prior information, and (c) omics annotation. The inclusion of this information in the analysis may enhance classification performance and feature selection, but is not straightforward in the standard regression setting. As a solution to this problem, we propose a group-regularized (logistic) elastic net regression method, where each penalty parameter corresponds to a group of features based on the external information. The method, termed \texttt{gren}, makes use of the Bayesian formulation of logistic elastic net regression to estimate both the model and penalty parameters in an approximate empirical-variational Bayes framework. Simulations and an application to a colon cancer microRNA study show that, if the partitioning of the features is informative, classification performance and feature selection are indeed enhanced.}
	\end{abstract}

	\noindent\textbf{Keywords}: Empirical Bayes; High-dimensional data; Prediction; Variational Bayes

	\noindent\textbf{Software available from}: \url{https://CRAN.R-project.org/package=gren}

	\section{Introduction}
	Prediction from high-dimensional data is a challenge common to many fields of research. Examples of such high-dimensional prediction problems arise in computer vision, stock market prediction and disease diagnosis from omics data. In this paper, we specifically focus on the latter setting. High dimensionality of data introduces several issues in the estimation of prediction models, especially: (a) unidentifiable models, (b) highly correlated predictor variables and (c) non-trivial variable selection.

	Attempts to tackle one or more of these issues have lead to the development of many prediction methods \cite[]{fan_selective_2010}. Here we will focus on the penalization approach. A well-known and popular penalization method is elastic net regression \cite[]{zou_regularization_2005}, with its special cases ridge regression \cite[]{hoerl_ridge_1970} and lasso regression \cite[]{tibshirani_regression_1996}. The elastic net approach takes issues (a)-(c) into account and has yielded many successful extensions. Recent work on elastic net regression (and its special cases) has focussed on increasing prediction accuracy by the inclusion of prior knowledge on the variables \cite[]{lee_variable_2017,van_de_wiel_better_2016,tai_incorporating_2007}.

	In omics research, sources of such prior knowledge on the variables are often available. The information can, for example, come in the form of (a) results on the same molecular features obtained in a previous study (e.g. $p$-values), (b) information from a publicly available database that summarizes the prior information on the molecular features involved (e.g., the Cancer Gene Census \cite[]{futreal_census_2004}), and (c) omics annotation (e.g. the location of a gene on the chromosome). Although this information can rarely be directly included in the statistical analysis, it may still be useful and informative for the study at hand.

	An example of such a study with extra information available is investigated in Section \ref{sec:app}. INTRODUCE DATA. This problem is not straightforward to solve with the standard methods available in literature. At first sight, such an \textit{a priori} grouping of the features suggests the group lasso \cite[]{meier_group_2008, yuan_model_2006} (or one of its extensions), which penalises and selects features at the group level. There are two limitations to the group lasso: (a) it selects groups as a whole instead of single features and (b) it is not adaptive: the amount of penalization a group receives depends on a `global' penalty parameter common to all groups. Extensions such as the sparse group lasso \cite[]{simon_sparse-group_2013} partly deal with (a), but do not address (b).

	A natural way to deal with (b) is through differential penalization: that is, each group of variables receives its own penalty parameter. An apparent issue with this differential penalization is the estimation of the penalty parameters. Naive estimation may be done by cross-validation. However, cross-validation requires re-estimation of the model over a grid of penalty parameters. The size of this grid increases exponentially with the number of penalty parameters. Consequently, it quickly becomes computationally infeasible. We therefore propose an efficient alternative: empirical Bayes estimation of the penalty parameters, which corresponds to hyperparameter estimation in the Bayesian prior framework. Because of the ubiquity of binary outcome data in omics research, we focus on the logistic elastic net. Recently, \cite{zhang_novel_2018} introduced VBVS, a method similar to ours in spirit. Nevertheless, one clear advantage of our method over VBVS is the possibility of adaptive inclusion of external information on the features to aid in prediction and feature selection.

	In line with the above, our proposed method is (a) subjective: the few remaining tuning parameters may be cross-validated, (b) adaptive: prior information is automatically and subjectively weighed with respect to its informativeness, (c) fast compared to full Bayesian analysis or cross-validation, and (d) requires minimal input by the user: only the data and grouping of the features is required.

	We introduce the Bayesian logistic (generalised) elastic net model in Section \ref{sec:model}. In Section \ref{seq:estimation} we derive a variational approximation to this model and use this novel approximation in the empirical Bayes estimation of multiple, group-specific penalty parameters. In Sections \ref{sec:simulations}, and \ref{sec:app}, we demonstrate the approach in a simulation study and a real data example, respectively. In Section \ref{sec:discussion} we conclude by both discussing some differences and parallels between the proposed method, termed \texttt{gren}, and other methods in the literature.

@@

	Many forms of empirical Bayes exist, the most formal one being maximisation of the marginal likelihood with respect to the hyperparameters. The resulting hyperparameter estimates are then plugged into the prior. The marginal likelihood is often introduced as a measure of model evidence given the observed data and is computed by integrating the product of likelihood and prior with respect to the model parameters. In the case of the elastic net introduced in (\ref{eq:enmodel2}) this results in the following marginal empirical Bayes posterior for $\bbeta$:
	\begin{align}
	p_{\hat{\blambda}}(\bbeta | \y) & = \frac{\L (\y ; \bbeta) \pi_{\hat{\blambda}} (\bbeta) }{p_{\hat{\blambda}}(\y)} = \frac{\int_{\btau} \L (\y ; \bbeta) \pi_{\hat{\blambda}}(\bbeta | \btau) \pi_{\hat{\blambda}}(\btau) \, d\btau}{p_{\hat{\blambda}}(\y)}, \nonumber \\
	\hat{\blambda} & := \underset{\blambda}{\argmax} \log p_{\blambda} (\y) = \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} \L_{\blambda}(\y, \bbeta, \btau) \, d\bbeta d\btau \nonumber \\
	& = \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} \L (\y ; \bbeta) \pi_{\blambda}(\bbeta | \btau) \pi_{\blambda}(\btau) \, d\bbeta d\btau. \label{eq:mmlestimate}
	\end{align}
	The integrals in (\ref{eq:mmlestimate}) are intractable in the case of the elastic net. In the omics setting the integrals are also high-dimensional, in which case numerical and Monte Carlo approximation methods become tedious and computationally expensive. Moreover, Laplace approximation is known to suffer from low accuracy in many high-dimensional settings \cite[]{shun_laplace_1995}. In \cite{casella_empirical_2001} an EM algorithm is described that estimates the hyperparameters. This EM algorithm iteratively maximises the expected joint log likelihood, such that the sequence:
	\begin{equation}\label{eq:ebEM}
	\blambda^{(k + 1)} = \underset{\blambda}{\argmax} \E_{\bbeta, \btau | \y} \[ \log \L_{\blambda}(\y, \bbeta, \btau) | \blambda^{(k)} \]
	\end{equation}
	converges to a local maximum of the marginal likelihood. The difficulty herein is in the calculation of the expected joint log likelihood. \cite{casella_empirical_2001} suggests to approximate the expectation by its Monte Carlo expectation. Although elegant and simple, this method requires a converged MCMC sample from the posterior for every iteration: a computationally intensive procedure. \cite{roy_selection_2017} introduce generalised importance sampling for the Bayesian elastic net, such that we just need a limited, pre-specified number of MCMC chains. However, this still requires several converged chains, so is not feasible in many high-dimensional omics settings. We propose to tackle this problem by approximating the expectation in (\ref{eq:ebEM}) using variational Bayes.

@@

@@

	\section{Simulations}\label{sec:simulations}
	\subsection{Setup}\label{sec:setup}
	To assess variable selection and predictive performance we conduct a simulation study in which we compare \texttt{gren} to the regular elastic net and ridge models, \texttt{GRridge} \cite[]{van_de_wiel_better_2016}, sparse group lasso (\texttt{SGL}) \cite[]{simon_sparse-group_2013}, group mimimax concave penalty (\texttt{grMCP}) \cite[]{breheny_penalized_2009}, and the group exponential lasso (\texttt{gel}) \cite[]{breheny_group_2015}. \texttt{GRridge} is similar to \texttt{gren} in the sense that it estimates group-specific penalty multipliers. The two main differences with \texttt{gren} are (i) the absence of an $L_1$-norm penalty and (ii) the estimation procedure. The other methods are extensions of the group lasso \cite[]{meier_group_2008, yuan_model_2006} and not adaptive on the group level. In contrast to the original group lasso, all of them are able to select single features, instead of just on the group level.

	For the regular elastic net, \texttt{gren}, \texttt{SGL}, \texttt{grMCP}, and \texttt{gel}, we fix the proportion of $L_1$-norm penalty to three different values: $\alpha \in \{0.05, 0.5, 0.95 \}$. The first setting closely resembles the ridge setting, where $\alpha=0$, while the third setting is similar to the lasso with $\alpha=1$. For the methods that do automatic variable selection (all methods except the regular ridge and \texttt{GRridge}), we estimate the models for a range of model sizes.

	Predictive performance is measured by area under the receiver operator curve (AUC) and Brier skill score (BSS). AUC gives the area under the curve of sensitivity versus $1-$ specificity and equals the probability that the classifier ranks a randomly chosen case higher than a randomly chosen non-case. The BSS is a normalised version of the Brier score: $\text{BSS} = 1 - \frac{\sum_{i=1}^n (y_i - \hat{y}_i)^2}{\sum_{i=1}^n (y_i - \bar{y})^2}$,
	where $\hat{y}_i=\expit(\x_i \tr \hat{\bbeta})$ is the prediction for observation $i$ and $\bar{y}=n^{-1} \sum_{i=1}^n y_i$. The BSS measures the fraction of Brier score improvement compared to a non-informative model. Quality of feature selection is assessed by Cohen's kappa \cite[]{cohen_coefficient_1960} and parameter estimation by mean squared error (MSE). Both are presented in SM Section 10.

	We simulated data according to five different scenarios: 
	\begin{enumerate}
		\item differential signal between the groups and uniformly distributed model parameters,
		\item a large number of small groups of features,
		\item no differential signal between the groups, but strong correlations within groups of features, 
		\item differential signal between the groups and heavy-tailed distributed model parameters, and
		\item a very sparse setting with no signal in some of the groups.
	\end{enumerate}
	Full descriptions of the scenarios are given in Section 10 of the SM. In the first scenario, the distribution of the model parameters is far off from the assumed ridge/elastic net/lasso distributions, so we assume that all methods will underperform. In the second scenario, we expect \texttt{gren} and \textt{GRridge} to overfit due to the large number of hyperparameters to estimate. The rest of the methods are conjectured to perform better, due to a lack of overfitting. In scenario three, we expect all methods to perform badly due to the high correlation between the features. In scenario three

	In all scenarios we created a training set of $n=100$ samples to estimate the models and a test set of $n_{\text{test}}=1000$ samples on which we compute the performance measures. The features $\x_i$, $i=1, \dots, n$, are sampled from a $p$-dimensional Gaussian distribution centred at zero, where we set $p=1000$. We let the covariance matrix of the features vary over the scenarios. Likewise for the model parameters $\beta_j$, $j=1, \dots, p$. We simulate the outcome data $y_i$, $i=1, \dots, n$, from a binary logistic model: $y_i \sim \mathcal{B}(1, \expit (\x_i \tr \bbeta))$. To mitigate the influence of random variation in the simulations, we repeated every scenario 100 times and report estimated penalty parameters and LOESS smoothed averages of the performance measures for a range of model sizes.

	 The results of this simulation are presented in Figures \ref{fig:sim1_performance1_set1}-\ref{fig:sim1_performance1_set3}.
	\begin{figure}[!h]
		\centering
		\includegraphics[width=1\linewidth]{gren_sim1_res1_performance1_set1.png}
		\caption{LOESS fit of (a) Cohen's kappa, (b) MSE, (c) AUC, and (d) BSS for ridge, \texttt{GRridge}, elastic net, \texttt{gren}, \texttt{SGL}, \texttt{grMCP}, and \texttt{gel} with $\alpha=0.05$ based on simulation scenario 1.}
		\label{fig:sim1_performance1_set1}
	\end{figure}

	\begin{figure}[!h]
		\centering
		\includegraphics[width=1\linewidth]{gren_sim1_res1_performance1_set2.png}
		\caption{LOESS fit of (a) Cohen's kappa, (b) MSE, (c) AUC, and (d) BSS for ridge, \texttt{GRridge}, elastic net, \texttt{gren}, \texttt{SGL}, \texttt{grMCP}, and \texttt{gel} with $\alpha=0.5$ based on simulation scenario 1.}
		\label{fig:sim1_performance1_set2}
	\end{figure}

	\begin{figure}[!h]
		\centering
		\includegraphics[width=1\linewidth]{gren_sim1_res1_performance1_set3.png}
		\caption{LOESS fit of (a) Cohen's kappa, (b) MSE, (c) AUC, and (d) BSS for ridge, \texttt{GRridge}, elastic net, \texttt{gren}, \texttt{SGL}, \texttt{grMCP}, and \texttt{gel} with $\alpha=0.95$ based on simulation scenario 1.}
		\label{fig:sim1_performance1_set3}
	\end{figure}

	The results of this simulation 1 are presented in Figures \ref{fig:sim4_performance1_set1}-\ref{fig:sim4_performance1_set3}.
	\begin{figure}[!h]
		\centering
		\includegraphics[width=1\linewidth]{gren_sim4_res1_performance1_set1.png}
		\caption{LOESS fit of (a) Cohen's kappa, (b) MSE, (c) AUC, and (d) BSS for ridge, \texttt{GRridge}, elastic net, \texttt{gren}, \texttt{SGL}, \texttt{grMCP}, and \texttt{gel} with $\alpha=0.05$ based on simulation scenario 4.}
		\label{fig:sim4_performance1_set1}
	\end{figure}

	\begin{figure}[!h]
		\centering
		\includegraphics[width=1\linewidth]{gren_sim4_res1_performance1_set2.png}
		\caption{LOESS fit of (a) Cohen's kappa, (b) MSE, (c) AUC, and (d) BSS for ridge, \texttt{GRridge}, elastic net, \texttt{gren}, \texttt{SGL}, \texttt{grMCP}, and \texttt{gel} with $\alpha=0.5$ based on simulation scenario 4.}
		\label{fig:sim4_performance1_set2}
	\end{figure}

	\begin{figure}[!h]
		\centering
		\includegraphics[width=1\linewidth]{gren_sim4_res1_performance1_set3.png}
		\caption{LOESS fit of (a) Cohen's kappa, (b) MSE, (c) AUC, and (d) BSS for ridge, \texttt{GRridge}, elastic net, \texttt{gren}, \texttt{SGL}, \texttt{grMCP}, and \texttt{gel} with $\alpha=0.95$ based on simulation scenario 4.}
		\label{fig:sim4_performance1_set3}
	\end{figure}

	\subsection{Results}\label{sec:results}

	\section{Application}\label{sec:app}

>>>>

\documentclass[oupdraft]{bio}
% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
\usepackage{url,bm,algorithm,algpseudocode,mathtools,bbm,pgfplotstable}

@@

@@

	% Title of paper
	\title{Adaptive group-regularized logistic elastic net regression}

	% List of authors, with corresponding author marked by asterisk
	\author{MAGNUS M. M\"UNCH$^{\ast 1,2}$, CAREL F.W. PEETERS$^{1}$, AAD W. VAN DER VAART$^{2}$, \\ MARK A. VAN DE WIEL$^{1,3}$\\[4pt]
		% Author addresses
		\textit{1. Department of Epidemiology \& Biostatistics, Amsterdam Public Health research institute, VU University Medical Center, PO Box 7057, 1007 MB Amsterdam, The Netherlands\\
			2. Mathematical Institute, Leiden University, Leiden, The Netherlands\\
			3. Department of Mathematics, VU University, Amsterdam, The Netherlands}
		\\[2pt]
		% E-mail address for correspondence
		{m.munch@vumc.nl}}

	% Running headers of paper:
	\markboth%
	% First field is the short list of authors
	{M. M. M\"unch and others}
	% Second field is the short title of the paper
	{Adaptive group-regularized logistic elastic net regression}

@@

	% Add a footnote for the corresponding author if one has been
	% identified in the author list
	\footnotetext{To whom correspondence should be addressed.}

	\begin{abstract}
		{In high-dimensional data settings, additional information on the features is often available. Examples of such external information in omics research are: (a) $p$-values from a previous study, (b) a summary of prior information, and (c) omics annotation. The inclusion of this information in the analysis may enhance classification performance and feature selection, but is not straightforward in the standard regression setting. As a solution to this problem, we propose a group-regularized (logistic) elastic net regression method, where each penalty parameter corresponds to a group of features based on the external information. The method, termed \texttt{gren}, makes use of the Bayesian formulation of logistic elastic net regression to estimate both the model and penalty parameters in an approximate empirical-variational Bayes framework. Simulations and an application to a colon cancer microRNA study show that, if the partitioning of the features is informative, classification performance and feature selection are indeed enhanced.}
		{Empirical Bayes; High-dimensional data; Prediction; Variational Bayes}
	\end{abstract}

	\section{Introduction}
	Prediction from high-dimensional data is a challenge common to many fields of research. Examples of such high-dimensional prediction problems arise in computer vision, stock market prediction and disease diagnosis from omics data. In this paper, we specifically focus on the latter setting. High dimensionality of data introduces several issues in the estimation of prediction models, especially: (a) unidentifiable models, (b) highly correlated predictor variables and (c) non-trivial selection of variables.

	Attempts to tackle one or more of these issues have lead to the development of many prediction methods \cite[]{fan_selective_2010}. Here we will focus on the penalization approach. A well-known and popular penalization method is elastic net regression \cite[]{zou_regularization_2005}, with its special cases ridge regression \cite[]{hoerl_ridge_1970} and lasso regression \cite[]{tibshirani_regression_1996}. The elastic net approach takes issues (a)-(c) into account and has yielded many successful extensions. Recent work on elastic net regression (and its special cases) has focussed on increasing prediction accuracy by the inclusion of prior knowledge on the variables \cite[]{van_de_wiel_better_2016,tai_incorporating_2007}.

	In omics research, sources of such prior knowledge on the variables are often available. The information can, for example, come in the form of (a) results on the same molecular features obtained in a previous study (e.g. $p$-values), (b) information from a publicly available database that summarizes the prior information on the molecular features involved (e.g., the Cancer Gene Census \cite[]{futreal_census_2004}), (c) omics annotation (e.g. the location of a gene on the chromosome) and (d) response-independent summary statistics, such as sample standard deviation of the features. Although this information can rarely be directly included in the statistical analysis, it may still be useful and informative for the study at hand.

	A natural way of including these information types is through differential penalization: that is, each group of variables receives its own penalty parameter. An apparent issue with this differential penalization is the estimation of the penalty parameters. Naive estimation may be done by cross-validation. However, cross-validation requires re-estimation of the model over a grid of penalty parameters. The size of this grid increases exponentially with the number of penalty parameters. Consequently, it quickly becomes computationally infeasible. We therefore propose an efficient alternative: empirical Bayes estimation of the penalty parameters, which corresponds to hyperparameter estimation in the Bayesian prior framework. Because of the ubiquity of binary outcome data in omics research, we focus on the logistic elastic net.

	We introduce the Bayesian logistic (generalised) elastic net model in Section \ref{sec:model}. In Section \ref{seq:estimation} we derive a variational approximation to this model and use this novel approximation in the empirical Bayes estimation of multiple, group-specific penalty parameters. In Section \ref{sec:extensions} we introduce some extensions of the method, including, among others, an extension to unpenalized covariates. In Sections \ref{sec:simulations}, and \ref{sec:app}, we demonstrate the approach in a simulation study and a microRNA colon cancer data example, respectively. In Section \ref{sec:discussion} we conclude by both discussing some differences and parallels between the proposed method, termed \texttt{gren}, and other methods in the literature.

@@

	Many forms of empirical Bayes exist, the most formal one being maximisation of the marginal likelihood with respect to the hyperparameters. The resulting hyperparameter estimates are then plugged into the prior. The marginal likelihood is often introduced as a measure of model evidence given the observed data and is computed by integrating the product of likelihood and prior with respect to the model parameters. In the case of the elastic net introduced in (\ref{eq:enmodel2}) this results in the following marginal empirical Bayes posterior for $\bbeta$:
	\begin{align}
	p_{\hat{\blambda}}(\bbeta | \y) & = \frac{\L (\y ; \bbeta) \pi_{\hat{\blambda}} (\bbeta) }{p_{\hat{\blambda}}(\y)} = \frac{\int_{\btau} \L (\y ; \bbeta) \pi_{\hat{\blambda}}(\bbeta | \btau) \pi_{\hat{\blambda}}(\btau) \, d\btau}{p_{\hat{\blambda}}(\y)}, \nonumber \\
	\hat{\blambda} & := \underset{\blambda}{\argmax} \log p_{\blambda} (\y) = \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} \L_{\blambda}(\y, \bbeta, \btau) \, d\bbeta d\btau \nonumber \\
	& = \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} \L (\y ; \bbeta) \pi_{\blambda}(\bbeta | \btau) \pi_{\blambda}(\btau) \, d\bbeta d\btau. \label{eq:mmlestimate}
	\end{align}
	The integrals in (\ref{eq:mmlestimate}) are intractable in the case of the elastic net. In the omics setting the integrals are also high-dimensional, in which case numerical and Monte Carlo approximation methods become tedious and computationally expensive. Moreover, Laplace approximation is known to suffer from low accuracy in many high-dimensional settings \cite[]{shun_laplace_1995}. In \cite{casella_empirical_2001} an EM algorithm is described that estimates the hyperparameters. This EM algorithm iteratively maximises the expected joint log likelihood, such that the sequence:
	\begin{equation}\label{eq:ebEM}
	\blambda^{(k + 1)} = \underset{\blambda}{\argmax} \E_{\bbeta, \btau | \y} \[ \log \L_{\blambda}(\y, \bbeta, \btau) | \blambda^{(k)} \]
	\end{equation}
	converges to a local maximum of the marginal likelihood. The difficulty herein is in the calculation of the expected joint log likelihood. \cite{casella_empirical_2001} suggests to approximate the expectation by its Monte Carlo expectation. Although elegant and simple, this method requires a converged MCMC sample from the posterior for every iteration: a computationally intensive procedure. We propose to tackle this problem by approximating the expectation in (\ref{eq:ebEM}) using variational Bayes.

@@

	\section{Extensions}\label{sec:extensions}
	\subsection{Unpenalized covariates}
	Inclusion of an intercept $\beta_0$ is achieved by appending the data matrix $\X$ with a column of ones. Penalization of such an intercept parameter is not desirable \cite[]{Hastie_Elements_2009}. Additionally, it is often desirable to include unpenalized covariates in the model. Common examples of such  covariates in clinical research are patient characteristics, such as age, BMI, and sex. In the following we assume the intercept to be included in the unpenalized covariates.

	To include unpenalized covariates, we divide the model matrix into two parts $\X = \begin{bmatrix}\X_u & \X_r\end{bmatrix}$, where $\X_u$ are the $u$ unpenalized variables and $\X_r$ are the $r$ penalized variables. Let $\bLambda'_*$ and $\Z_*$ be the matrices $\bLambda'$ and $\Z$ prepended with $u$ zero columns and $u$ zero rows. Then we have for the current estimate of the covariance matrix $\bSigma$:
	\begin{align}\label{eq:unpencov}
	\bSigma &=\(\X \tr \bOmega \X + \lambda_2 \bLambda'_* + \frac{\lambda_1 \sqrt{\lambda_2}}{2} \bLambda'_* \Z_*\)^{-1} \nonumber \\
	& = 
	\begin{bmatrix} 
	\X_u \tr \bOmega \X_u & \X_u \tr \bOmega \X_r \\
	\X_r \tr \bOmega \X_u & \X_r \tr \bOmega \X_r + \lambda_2 \bLambda' + \frac{\lambda_1 \sqrt{\lambda_2}}{2} \bLambda' \Z
	\end{bmatrix}^{-1}.
	\end{align}
	With the choice of blocks as in (\ref{eq:unpencov}), blockwise inversion renders the largest required matrix inverse $(\X_r \tr \bOmega \X_r + \lambda_2 \bLambda' + \frac{\lambda_1 \sqrt{\lambda_2}}{2} \bLambda' \Z)^{-1}$, of dimension $p \times p$. Inversion of this matrix is done efficiently by applying the Woodbury identity.

	\subsection{Monotonicity of the penalty parameters}
	Enforcing monotonicity of the penalty multipliers is desirable in some settings. An example of such a setting is if we have $p$-values from a previous, related study available. A $p$-values-based partitioning of the variables may, \textit{a priori}, be expected to render penalty multipliers that increase monotonically with $p$-value. That is, larger $p$-values may be expected to yield at least as large penalty multipliers as variables with smaller $p$-values. We propose to include this \textit{a priori} assumption by requiring the penalty multipliers to increase monotonically with $p$-value, thereby also stabilizing their estimates.

	A natural way of enforcing monotonicity is to extend the constraint in (\ref{eq:mmlupdateequation}) with $\lambda'_1 \leq \dots \leq \lambda'_G$. The problem is still convex and may be numerically solved. From experience, however, we note that in combination with this constrained optimisation, the EM algorithm described in Section~\ref{seq:empiricalbayes} often converges to a local optimum close to the initialisation. We therefore enforce monotonicity through a \textit{post hoc} isotonic regression on the penalty multipliers after every optimisation step.

@@

	\subsection{Multiple partitions}
	In many cases the features may be partitioned in more than one way. For example, we may have both information on annotation and $p$-values from a previous study available. A naive way of incorporating multiple partitions is to cross-tabulate the partitions and create a separate group for every combination. This poses two problems: (i) The number of penalty parameters increases exponentially with the number of partitions and (ii) some of these combinations may contain only few features, so that the estimation procedure becomes unstable. We propose to stabilise the procedure and keep the number of parameters to estimate manageable by modelling the penalty parameters multiplicatively.

	We describe our implementation here for two partitions of the features. To this end, let $(\G_1(1), \dots, \G_1(G_1))$ and $(\G_2(1), \dots, \G_2(G_2))$ denote the two partitions, containing $G_1$ and $G_2$ groups respectively. Furthermore, in the following we assume that the empty sum and empty product evaluate to 0 and 1, respectively. In this two-partition setting we have two penalty multipliers per feature, represented by $\lambda'_{g_1}$ and $\lambda''_{g_2}$, respectively. The generalised frequentist elastic net estimator and corresponding conditional prior are now:
	\begin{subequations}\label{eq:multpart}
		\begin{align}
		\hat{\bbeta} &:= \underset{\bbeta}{\argmax} \ell(\y ; \bbeta) -  \frac{\lambda_1}{2} \sum_{g_1=1}^{G_1} \sum_{g_2=1}^{G_2} \sqrt{\lambda'_{g_1} \lambda''_{g_2}} \sum_{\mathclap{\substack{j \in \G_1(g_1) \\ \cap \G_2(g_2)}}} | \beta_j| - \frac{\lambda_2}{2} \sum_{g_1=1}^{G_1} \sum_{g_2=1}^{G_2} \lambda'_{g_1} \lambda''_{g_2} \sum_{\mathclap{\substack{j \in \G_1(g_1) \\ \cap \G_2(g_2)}}} \beta_j^2, \\
		\bbeta | \btau & \sim \prod_{g_1=1}^{G_1} \prod_{g_2=1}^{G_2} \prod_{\substack{j \in \G_1(g_1) \\ \cap \G_2(g_2)}} \mathcal{N} \left(0,\frac{1}{ \lambda'_{g_1} \lambda''_{g_2} \lambda_2} \frac{\tau_j - 1}{\tau_j} \right).
		\end{align}
	\end{subequations}
	After switching to the parametrisation in \cite{friedman_regularization_2010}, the new penalty multiplier estimates $\blambda'= \begin{bmatrix} \lambda'_1 & \cdots & \lambda'_{G_1} \end{bmatrix} \tr, \blambda''= \begin{bmatrix} \lambda''_1 & \cdots & \lambda''_{G_2} \end{bmatrix} \tr$ are given by:
	\begin{subequations}\label{eq:estmultpart}
		\begin{alignat}{2}
		\blambda'^{(k+1)}, \blambda''^{(k+1)} &:= \underset{\blambda', \blambda''}{\argmax} && \Bigg\{ \frac{1}{2} \sum_{g_1=1}^{G_1} |\G_1(g_1)| \log (\lambda'_{g_1}) + \frac{1}{2} \sum_{g_2=1}^{G_2} |\G_2(g_2)| \log (\lambda''_{g_2}) \\ & && - \frac{(1 - \alpha) \lambda}{4} \sum_{g_1=1}^{G_1} \sum_{g_2=1}^{G_2} \lambda'_{g_1} \lambda''_{g_2} d^{(k)}_{g_1 g_2} \Bigg\} \nonumber \\       & \text{subject to } && \prod_{g_1=1}^{G_1} \prod_{g_2=1}^{G_2} (\lambda'_{g_1} \lambda''_{g_2})^{|\G_1(g_1) \cap \G_2(g_2)|} = 1,
		\end{alignat}
	\end{subequations}
	with the $d_{g_1 g_2}$ terms calculated as $\sum_{\substack{j \in \G_1(g_1) \\ \cap \G_2(g_2)}} \[\bSigma^{(k)}_{jj} + (\bmu^{(k)}_j)^2\]\(1 + \alpha \lambda^{1.5} \sqrt{\frac{1 - \alpha}{8 \chi_j^{(k)}}}\)$. The optimisation in (\ref{eq:estmultpart}) is again a convex problem that is easily solved by some numerical optimisation routine. This method naturally generalises to more than two partitions of the data.

	In (\ref{eq:multpart}) we model the two partition-specific penalty multipliers in a multiplicative way, mainly for convenience. First, there are computational reasons: The separation of the square root in the $L_1$-norm penalty term into two terms facilitates the numerical estimation. Additionally, multiplicative modelling of the partitions allows for more flexibility in the estimates. To see this consider the obvious alternative: additive partition-specific multipliers. Since the penalty multipliers are strictly positive, the constraint on the geometric mean makes one large additive penalty estimate difficult to compensate with a smaller estimate, thereby impairing flexibility of the model. Large multiplicative penalty multipliers are easier to compensate for by smaller penalty multipliers, rendering this the less restrictive option for modelling the partitions.

	\section{Simulations}\label{sec:simulations}
	\subsection{Setup}\label{sec:setup}
	To assess variable selection and predictive performance we conduct a simulation study in which we compare \texttt{gren} to the regular elastic net and ridge models, and \texttt{GRridge} \cite[]{van_de_wiel_better_2016}. \texttt{GRridge} is similar to \texttt{gren} in the sense that it estimates group-specific penalty multipliers. The two main differences with \texttt{gren} are (i) the absence of an $L_1$-norm penalty and (ii) the estimation procedure. For the regular elastic net and \texttt{gren}, we fix the proportion of $L_1$-norm penalty to three different values: $\alpha \in \{0.05, 0.5, 0.95 \}$. The first setting closely resembles the ridge setting, where $\alpha=0$, the third one is similar to the lasso with $\alpha=1$. For the methods that do automatic variable selection (the regular elastic net and \texttt{gren}), we estimate the models for a range of model sizes. Likewise for \texttt{GRridge}, where we use the \textit{post hoc} variable selection method described by \cite{novianti_better_2017}.

	The quality of feature selection is assessed by Cohen's kappa \cite[]{cohen_coefficient_1960}. This kappa is measures similarity between two sets of categorical values. One set contains the feature selection indicators: $s(j) := \mathbbm{1}\{\hat{\beta}_j \neq 0\}$, while the other set contains the true non-zero feature indicators: $t(j) := \mathbbm{1}\{\beta_j \neq 0\}$. We calculate Cohen's $\kappa$ as:
	\begin{align*}
	\kappa & = \frac{f_o - f_e}{1 - f_e}, \text{ where } f_o = p^{-1}\sum_{j=1}^p \mathbbm{1}\{s(j)=t(j)\} \\ & \text{ and } f_e = p^{-2} \left\{ \(\sum_{j=1}^p s(j)\) \cdot \(\sum_{j=1}^p t(j)\) + \[\sum_{j=1}^p (1 - s(j)) \] \cdot \[\sum_{j=1}^p (1 - t(j)) \]\right\}.
	\end{align*}
	Here, $f_o$ and $f_e$ denote the frequencies of correctly identified and expected correctly identified features, respectively. A positive or negative kappa indicate better and worse feature selection than expected by chance, respectively. In addition, we evaluate estimation accuracy by mean squared error: $\text{MSE} = p^{-1} \sum_{j=1}^p (\beta_j - \hat{\beta}_j)^2$.

	Predictive performance is measured by area under the receiver operator curve (AUC) and Brier skill score (BSS). AUC gives the area under the curve of sensitivity versus $1-$ specificity and equals the probability that the classifier ranks a randomly chosen case higher than a randomly chosen non-case. The BSS is a normalised version of the Brier score: $\text{BSS} = 1 - \frac{\sum_{i=1}^n (y_i - \hat{y}_i)^2}{\sum_{i=1}^n (y_i - \bar{y})^2}$,
	where $\hat{y}_i=\expit(\x_i \tr \hat{\bbeta})$ is the prediction for observation $i$ and $\bar{y}=n^{-1} \sum_{i=1}^n y_i$. The BSS measures the fraction of Brier score improvement compared to a non-informative model.

	We created a training set of $n=100$ samples to estimate the models and a test set of $n_{\text{test}}=1000$ samples on which we compute the performance measures. The features $\x_i$, $i=1, \dots, n$, are sampled from a $p$-dimensional Gaussian distribution centred at zero, where we set $p=1000$. We introduce correlation between the features through a block diagonal covariance matrix for the $\x_i$. The blocks are of sizes $25 \times 25$, with off-diagonals set to $\rho=0.7$ and diagonals to $\sigma^2 = 1$.

	The model parameters $\beta_j$, $j=1, \dots, p$, are simulated in four groups of 250 with differing signal strengths and are created in two steps. First, we draw 250 parameters from the elastic net distribution, parametrised as in \cite{friedman_regularization_2010}, where $\alpha=0.5$ and $\lambda=100$ in all groups, and different penalty multipliers, $\lambda' \in \{ 0.14, 0.51, 1.95, 7.39 \}$. The second step consists of setting the (in absolute value) smallest 125 parameters in a group to zero. This results in a total of 500 zero and 500 non-zero parameters, evenly distributed over the four groups.
